Models

### 2021-10-18 analysis/05_model_metrics.Rmd
Add interaction models to get contrasts between levels and within level slopes.

model_data <- metrics %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
                       "FRic", "FEve", "FDiv", "S"),
         !rarefyID %in% missing_null) %>%
  group_by(metric) %>%
  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
  mutate(year_scaled = scale(year), 
         scale_center = attributes(year_scaled)$`scaled:center`, 
         scale = attributes(year_scaled)$`scaled:scale`) %>%
  ungroup() %>%
  #this study has many low species richness observations, so FD metrics weren't calculated, just drop it
  filter(study_id != 348)

@@

  fit <- data %>%
    filter(metric == metric_name) %>%
    mutate(year_scaled = scale(year)) %>%
    qlmer(formula = formula, data = ., ...)

  tidy(fit$result) %>%
   mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
}

#get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
```

Create dataframe of models to fit
```{r}
# raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
#                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
#                          taxa = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
#                          realm = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
#                          climate = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
# 
# log_value_mods <- tibble(metric_name = c("S"), 
#                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
#                          taxa = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
#                          realm = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
#                          climate = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
                         taxa = "value ~ year_scaled + as.factor(taxa) + (year_scaled|study_id/rarefyID)",
                         realm = "value ~ year_scaled + as.factor(realm) + (year_scaled|study_id/rarefyID)",
                         climate = "value ~ year_scaled + as.factor(climate) + (year_scaled|study_id/rarefyID)") %>%
  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

log_value_mods <- tibble(metric_name = c("S"), 
                         base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
                         taxa = "logvalue ~ year_scaled + as.factor(taxa) + (year_scaled|study_id/rarefyID)",
                         realm = "logvalue ~ year_scaled + as.factor(realm) + (year_scaled|study_id/rarefyID)",
                         climate = "logvalue ~ year_scaled + as.factor(climate) + (year_scaled|study_id/rarefyID)") %>%
  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

@@

#Turn of REML for these models 
bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)

#One model still won't converge
bad_models2 <- bad_mod_results %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(mods, by = c("metric" = "metric_name", "model_id"))

#Fit with a different algorithm
bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))

metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
          bad_mod_results %>% filter(is.na(warning)),
          bad_mod_results2 %>% filter(is.na(warning)))

usethis::use_data(metric_model_table)
pins::pin(metric_model_table, board = "github")
```
#### Process Model Output

@@

Pairwise correlation analysis
```{r}
library(psych)
#slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))

slope_corr_bootstrap <- corCi(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
```

Clustering 
```{r}
library(clValid)
scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)

scale_slopes <- slopes_wide %>%  
  drop_na() %>%
  select(study_id, starts_with("SES"), S, Jaccard_base) %>%
  column_to_rownames("study_id") %>%
  mutate(across(.fns = scale2)) %>%
  as.matrix()

clmethods <- c("hierarchical","kmeans","pam")
intern <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "internal")
stab <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "stability")

summary(intern)
optimalScores(stab)

#get the hierarchical clustering object
clust <- intern@clusterObjs$hierarchical
cat5 <- cutree(clust, k = 5)
cat3 <- cutree(clust, k = 3)

slopes_clust <- slopes_wide %>% 
  ungroup() %>%
  left_join(as.data.frame(cat3) %>%
  rownames_to_column("study_id")) %>%
  left_join(as.data.frame(cat5) %>%
  rownames_to_column("study_id"))
```

```{r}
slopes_clust %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S", "Jaccard_base") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }
```

####Figures

Pairwise plots of study-level slopes
```{r}
#pairwise plots using GGally
ggpairs(slopes_wide %>% 
          select(-c("study_id", "FDiv", "FEve", "FRic")))

#pairwise plots using GGPlot
gatherpairs <- function(data, ..., 
                        xkey = '.xkey', xvalue = '.xvalue',
                        ykey = '.ykey', yvalue = '.yvalue',
                        na.rm = FALSE, convert = FALSE, factor_key = FALSE) {
  vars <- quos(...)
  xkey <- enquo(xkey)
  xvalue <- enquo(xvalue)
  ykey <- enquo(ykey)
  yvalue <- enquo(yvalue)

  data %>% {
    cbind(gather(., key = !!xkey, value = !!xvalue, !!!vars,
                 na.rm = na.rm, convert = convert, factor_key = factor_key),
          select(., !!!vars)) 
  } %>% gather(., key = !!ykey, value = !!yvalue, !!!vars,
               na.rm = na.rm, convert = convert, factor_key = factor_key)
}

slopes_wide %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue)) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }

slopes_wide %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
  ggplot(aes(x = S, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-1.3, 1.3) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Richness Slope")

slopes_wide %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
  ggplot(aes(x = SES_FRic, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-.25, .25) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Functional Richness Slope")
```

>>>>

model_data <- metrics %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
                       "FRic", "FEve", "FDiv", "S",
                       "bodymass_value"),
         !rarefyID %in% missing_null) %>%
  group_by(metric) %>%
  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
  mutate(year_scaled = scale(year), 
         scale_center = attributes(year_scaled)$`scaled:center`, 
         scale = attributes(year_scaled)$`scaled:scale`) %>%
  ungroup() %>%
  #this study has many low species richness observations, so FD metrics weren't calculated, just drop it
  filter(study_id != 348)

@@

  fit_data <- data %>%
    filter(metric == metric_name) %>%
    mutate(year_scaled = scale(year),
           duration_scaled = scale(duration, scale = FALSE),
           startyear_scaled = scale(startyear, scale = FALSE),
           taxa = as.factor(taxa),
           realm = as.factor(realm),
           climate = as.factor(climate))

    fit <- qlmer(formula = formula, data = fit_data, ...)

  if(!model_id %in% c("base", "base_dur", "base_srt")){
    level_slopes <- get_fixed_slopes(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data) %>%
      mutate(effect = "group_slope", group = model_id) %>%
      rename(std.error = SE, estimate = year_scaled.trend, term = model_id)

    int_contrasts <- get_int_contrasts(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data) %>%
      mutate(group = model_id, effect = "contrast") %>%
      rename(std.error = SE, statistic = t.ratio, term = contrast) %>%
      select(-year_scaled)

    df <- tidy(fit$result) %>% 
      bind_rows(level_slopes, int_contrasts) %>%
      mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
  }else{
  tidy(fit$result) %>%
   mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
  }
}

get_int_contrasts <- function(metric, model, model_id, model_data){
  formula <- paste("pairwise ~", model_id, "| year_scaled")
  emm <- emmeans(model,  mode = "satterth", lmerTest.limit = 20105, specs = as.formula(formula), data = model_data)

  return(as_tibble(emm$contrasts))
}

get_fixed_slopes <- function(metric, model, model_id, model_data){
  formula <- paste("~", model_id)
  fit_trends <- emtrends(model, specs = as.formula(formula), var="year_scaled",  mode = "satterth", lmerTest.limit = 20105, data = model_data) %>% as_tibble()

  return(fit_trends)
}
#get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
```

Create dataframe of models to fit
```{r}
raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
                         taxa = "value ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
                         realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

log_value_mods <- tibble(metric_name = c("S"), 
                         base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
                         taxa = "logvalue ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
                         realm = "logvalue ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
                         climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

@@

#Turn off REML for these models, and they fit! 
bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)

metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
          bad_mod_results %>% filter(is.na(warning)))

usethis::use_data(metric_model_table)
pins::pin(metric_model_table, board = "github")
```

Models with interactions
```{r}
int_raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
                         base_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
                         base_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
  pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
               names_to = "model_id", values_to = "formula")

int_log_value_mods <- tibble(metric_name = c("S"),
                         base_dur = "logvalue ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
                         base_srt = "logvalue ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
  pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
               names_to = "model_id", values_to = "formula")

# mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
#   select(metric_name, formula, model_id)

#Fit models with base settings
int_mod_results <- pmap_dfr(bind_rows(int_raw_value_mods, int_log_value_mods), get_model, data = model_data)

#get the models that had convergence warnings
bad_models <- int_mod_results %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(int_raw_value_mods, by = c("metric" = "metric_name", "model_id")) 
```

#### Process Model Output

@@

### 2021-10-18 analysis/exporatory_scripts/problem_model.Rmd
Add interaction models to get contrasts between levels and within level slopes.

#get metadata
meta <- pins::pin_get("meta", board = "github") %>%
  select(study_id, realm, climate, habitat, protected_area, biome_map)

meta <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  select(study_id, rarefyid, taxa) %>% distinct() %>%
  left_join(meta)

#get metric data
metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  filter(!is.na(value)) %>%
  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  left_join(meta, c("rarefyID" = "rarefyid"))

#find rarefyID's that have inf value for some metrics and exclude
cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)

#get rarefyID's that don't have enough null samples
missing_null <- bind_rows(metrics %>% 
  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
  metrics %>% filter(n_missing_nulls > 100)) %>%
  pull(rarefyID) %>%
  unique()

model_data <- metrics %>% 
  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
  group_by(metric) %>%
  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
  mutate(year_scaled = scale(year), 
         scale_center = attributes(year_scaled)$`scaled:center`, 
         scale = attributes(year_scaled)$`scaled:scale`) %>%
  ungroup()
```

@@

Visualize distributions of response
```{r}
model_data %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
  ggplot() +
  geom_histogram(aes(x = logvalue),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of Log values")

model_data %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

Fit a single model to play around with 
```{r}
test_data <- model_data %>% 
  filter(metric == "FRic") %>%
  mutate(study_id = as.factor(study_id))

#lognormal
ln_fit <- lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), data = test_data,
               control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))

>>>>

model_data <- pins::pin_get("model-data", board = "github")
```

@@

Visualize distributions of response
```{r}
# model_data %>%
#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
#   ggplot() +
#   geom_histogram(aes(x = logvalue),
#                  bins = 60) +
#   facet_wrap(~metric, scales = "free") +
#   labs(title = "Distribution of Log values")
# 
# model_data %>%
#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
#   ggplot() +
#   geom_histogram(aes(x = value),
#                  bins = 60) +
#   facet_wrap(~metric, scales = "free") +
#   labs(title = "Distribution of raw values")
```

Fit a single model to play around with 
```{r}
test_data <- model_data %>% 
  filter(metric == "FRic") %>%
  mutate(year_scaled = scale(year),
           duration_scaled = scale(duration, scale = FALSE),
           startyear_scaled = scale(startyear, scale = FALSE),
         taxa = as.factor(taxa))
```

Different options for getting between group samples, fixed effects (test differences between intercept):
```{r}
library(emmeans)

#fixed effects contrasts
fit_fixed <- lmer(formula = "value ~ year_scaled*taxa + (year_scaled|study_id/rarefyID)", data = test_data)
m_emm <-  emmeans(fit_fixed,  mode = "satterth", lmerTest.limit = 20105, pairwise ~ taxa | year_scaled)

fit_cont <- as_tibble(m_emm$contrasts)
fit_trends <- emtrends(fit_fixed, ~ taxa, var="year_scaled",  mode = "satterth", lmerTest.limit = 20105) %>% as_tibble()

interactions::interact_plot(fit_fixed, pred = year_scaled, modx = taxa, plot.points = TRUE)
```

As a random effect:
```{r}
fit_rand <- lmer(formula = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)", data = test_data)
```

```{r}
#lognormal
ln_fit <- lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), data = test_data)
